---
title: "Practical Machine Learning Course Project"
author: "Anastasiia"
date: "15 november 2015 "
output: html_document
---
###Background

Using devices such as Jawbone Up, Nike FuelBand, and Fitbit it is now possible to collect a large amount of data about personal activity relatively inexpensively. These type of devices are part of the quantified self movement ??? a group of enthusiasts who take measurements about themselves regularly to improve their health, to find patterns in their behavior, or because they are tech geeks. One thing that people regularly do is quantify how much of a particular activity they do, but they rarely quantify how well they do it. In this project, your goal will be to use data from accelerometers on the belt, forearm, arm, and dumbell of 6 participants. They were asked to perform barbell lifts correctly and incorrectly in 5 different ways. More information is available from the website here: <http://groupware.les.inf.puc-rio.br/har> (see the section on the Weight Lifting Exercise Dataset). 

###Data 

The training data for this project are available here: 
<https://d396qusza40orc.cloudfront.net/predmachlearn/pml-training.csv>

The test data are available here: 
<https://d396qusza40orc.cloudfront.net/predmachlearn/pml-testing.csv>

###Objective

The goal of your project is to predict the manner in which they did the exercise. This is the "classe" variable in the training set. The analysis develops prediction model and uses it to predict 20 different test cases in the test data. 

###Data Loading
```{r, cache=TRUE}
setwd("/Users/anastasia/Documents/Coursera/Practical Machine Learning")
if (!file.exists("training.csv")) {download.file("http://d396qusza40orc.cloudfront.net/predmachlearn/pml-training.csv", destfile = "training.csv")}
if (!file.exists("test.csv")) {download.file("http://d396qusza40orc.cloudfront.net/predmachlearn/pml-testing.csv", destfile = "test.csv")}
trainData<- read.csv("training.csv", na.strings = c("NA", "","#DIV/0!"))
testData<- read.csv("test.csv", na.strings = c("NA", "","#DIV/0!"))
```

###Data Processing
```{r, cache=TRUE}
colTr  <- colnames(trainData)
colTes <- colnames(testData)
# Verify that the column names (excluding the last columns classe and problem_id) are identical in the training and test set
print(all.equal(colTr[1:length(colTr)-1], colTes[1:length(colTes)-1]))
```
```{r, cache=TRUE}
# Drop the first 7 columns as they are unnecessary for predicting
trainData<- trainData[,8:160]
testData <- testData[,8:160]
# Drop the rows if they contain more than 50% of NA values
trainData<-trainData[, colSums(is.na(trainData)) < nrow(trainData) * 0.5]
testData <-testData[, colSums(is.na(testData)) < nrow(testData) * 0.5]
```

###Data Splitting
There are 19,622 observations in the training set, so in order to reduce time and to be able to perform cross-validation, a training subset is created with 60% of the original training data set to be used for training and the remaining 40% to be used as the testing set (before final testing is performed).
```{r, cache=TRUE}
library(caret)
library(lattice)
library(ggplot2)
set.seed(777)
#str(trainData)
index <- createDataPartition(y=trainData$classe, p=0.6, list=FALSE)
train1 <- trainData[index, ]
test1  <- trainData[-index, ]
```

###Data fitting and predicting. Decision Tree.

```{r, cache=TRUE}
set.seed(777)
library(rpart)
# Grow tree 
fit <- rpart(classe ~ .,method="class", data=train1)
```

```{r, cache=TRUE}
#summary(fit) # detailed summary of splits
library(rpart.plot)
#plot the tree
prp(fit)
```

```{r}
set.seed(777)
library(caret)
c<-confusionMatrix(predict(fit,test1,type="class"), test1$classe)
# print out of sample error
print(c, digits=4)
print(1-c[["overall"]][["Accuracy"]])
```
###Data fitting and predicting. Random Forest.
```{r, cache=TRUE}
set.seed(777)
library(randomForest)
library(doParallel)
library(doMC)
registerDoMC(cores = 2)
# use parallel calculation for random forest fitting
train_control <- trainControl(method="cv", number=3)
model <- train(classe~., data=train1, method="rf", trControl = train_control)
print(model)
# show important features
varImp(model)
c<-confusionMatrix(predict(model,test1[,1:length(test1)-1],type="raw"), test1$classe)
print(c, digits=4)

print("Out Sample Error Rate")
print(1-c[["overall"]][["Accuracy"]])

print("In Sample Error Rate")
pred3<-predict(model,train1[,1:length(train1)-1],type="raw")
c<-confusionMatrix(predict(model,train1[,1:length(train1)-1],type="raw"), train1$classe)
print(1-c[["overall"]][["Accuracy"]])
```

###Evaluation

```{r, cache=TRUE}
predFinal <- predict(model, testData, type = "raw")
print(predFinal)
```

###Files for submitting

```{r, cache=TRUE}
for (i in seq(20)){
        fileName<-paste("problem",i,".txt",sep="_")
        write.table(predFinal[i],file=fileName,quote=FALSE,row.names=FALSE,col.names=FALSE)
}
```